---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document: 
    keep_md: true
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(#fig.width=10, fig.height=8, 
                      echo=FALSE, warning=FALSE, message=FALSE)
```

```{r}
#preamble

Sys.setlocale("LC_TIME", "English") #to have english output and labels

# load R-libraries
# ================
library(zoo) #time series library
library(timeDate)

```



## Loading and preprocessing the data

1. Load the data (i.e. read.csv())


The data of the activity tracker can be loaded with this code:

```{r, echo=TRUE}
dat<- read.table("activity.csv", header=T, sep=",")
str(dat) #check correct import

```


2. Process/transform the data (if necessary) into a format suitable for your
analysis

The data are conveniently transformed into a zoo time-series object. This is best done via a function "get.time()":

```{r, echo=TRUE}
get.time <- function(activity){
#re-format data into time series
#--------------        
        
        
        #extracts POSIX time from activity.csv data frame
        time.character <- paste(activity$date, formatC(activity$interval, width = 4, 
                                                       format = "d", flag = "0"), sep="-")
        #returns "e.g. "2012-10-01-0000"
        time.POSIXct <- strptime(time.character, format="%Y-%m-%d-%H%M", tz="GMT")
        return(time.POSIXct)
}

```


This is used to import the data into the zoo() object, which facilitates times series manipulation and aggregation.

```{r, echo=TRUE}
time.vec        <- get.time(dat)
steps   <- zoo(dat$steps, order.by=time.vec)
```

They look like this. Periods with "NA" values are labelled red.

```{r}
plot(steps, xlab="")
abline(v=as.numeric(index(steps[is.na(steps)])), col=2, lty=2)
axis(1, at=as.numeric(seq(index(steps[1]), index(steps[length(steps)]), by="1 week")), labels=F, tick=T)
````



## What is mean total number of steps taken per day?

1. Make a histogram of the total number of steps taken each day

```{r}

m1 <- aggregate(steps, by=as.Date(index(steps)), sum, na.rm=T)
m2 <- aggregate(steps, by=as.Date(index(steps)), mean, na.rm=T)
m3 <- aggregate(steps, by=as.Date(index(steps)), median, na.rm=T)


hist(m1, main="Histogram of total number of steps taken each day", xlab="steps/day")

```

2. Calculate and report the mean and median total number of steps taken per day

The individual took on average `r round(mean(m1, na.rm=T),0)` steps per day with a median of `r formatC(median(m1, na.rm=T),format="d")` steps.

## What is the average daily activity pattern?

The average daily activity pattern is nicely illustrated with  a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis). 

```{r}
m3 <- aggregate(steps, by=dat$interval, mean, na.rm=T)
plot(m3)

id.x.max <- which(m3==(max(m3)))
x.max     <-index(m3[id.x.max])
y.max <- as.numeric(m3[id.x.max])

points(x.max,y.max, col=2, lwd=2)
text(x=1.1*x.max, 0.98*y.max,pos=4, paste("max.steps =", round(y.max,0)),col=2 )
```

The 5-minute interval at `r x.max`, contains the maximum number of steps (`r y.max`) on average across all the days in the dataset.


## Imputing missing values

In total, the dataset contains `r sum(is.na(steps))` missing values.


A simple strategy for filling in all of the missing values in the dataset is to use the mean for that 5-minute interval as the best estimate. This can be done in the following way:

```{r, echo=TRUE}
#get the IDs of the missing values

id.na <- 1:length(steps)
id.na <- id.na[is.na(steps)]

#look for the IDs of the matching interval
id.matching.interval <- match(dat$interval[id.na], index(m3))

```

Then, a new dataset "steps.clean" can be created that is equal to the original dataset but with the missing data filled in.

```{r, echo=TRUE}
steps.clean <- steps
steps.clean[id.na] <- as.numeric(m3[id.matching.interval])
```


A new histogram of the total number of steps taken each day reveals a change in patterns. 

```{r}
m4 <- aggregate(steps.clean, by=as.Date(index(steps.clean)), sum, na.rm=T)
hist(m4, main="Histogram of total number of steps taken each day", xlab="steps/day")

#plot(m1)
#lines(m4, col=3)

d.mean <- mean(m4)-mean(m1, na.rm=T)
d.median <- median(m4)- median(m1, na.rm=T)
```

 
Now, the mean is `r formatC(mean(m4, na.rm=T),format="d")` and median `r formatC(median(m4), format="d")`. These values differ from the estimates from the first part of the assignment by `r round(d.mean/mean(m1, na.rm=T)*100, 0)` percent (`r round(d.median/median(m1, na.rm=T)*100,0)` for the median). As expected, imputing missing data increases the estimates of the total daily number of steps.


## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday"
and "weekend" indicating whether a given date is a weekday or weekend
day.

```{r, echo=TRUE}

dat$daytype <- ifelse(isWeekday(index(steps)), "Weekday", "Weekend")

id.weekday <- which(dat$daytype =="Weekday") #ID's of weekdays
```

2. Make a panel plot containing a time series plot (i.e. type = "l") of the
5-minute interval (x-axis) and the average number of steps taken, averaged
across all weekday days or weekend days (y-axis).

```{r}

m5 <- aggregate(steps.clean[id.weekday], by=dat$interval[id.weekday], mean)
m6 <- aggregate(steps.clean[-id.weekday], by=dat$interval[-id.weekday], mean)

plot(cbind(m5, m6), xlab="Time interval", ylab="steps", ylim=range(0:y.max),
     main="Step patterns of Weekdays (top) and Weekends (bottom)")
```

Yes, on weekends there are generally fewer steps and the activity pattern shows less activity in the morning and late afternoon. Instead, activity increases in afternoon >15:00 and in the evening around 20:00.